
---
unit_title: "Google sheets"
theme: wrangle
needs: [parse-basics]
---


```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Use googlesheets by Jenny Bryan to (suprise!) extract data from googlesheets.  Sheets is a surprisingly useful way of collaboratively collecting data (especially with google forms); googlesheets makes it easy to get that data into R and make use of it.

If you haven't already, install it:

```{r, eval = FALSE}
install.packages("googlesheets")
```

```{r setup}
library(googlesheets)
```

## Sheet sources

Two basic sources of sheets:

1. Your sheets
2. Public sheets

### Public sheets

<https://docs.google.com/spreadsheets/d/1twq7WtX5mor07gR8hjZ0bErOEP8iZJFxD6Q3PUSQ-Z8/pubhtml>

```{r}
gs_url("https://docs.google.com/spreadsheets/d/1twq7WtX5mor07gR8hjZ0bErOEP8iZJFxD6Q3PUSQ-Z8/pubhtml")
gs_read()
```

### Your sheets

Key challenge is identifying the sheet you want:

```{r}
gs_ls()
gs_ls("Gapminder")
```

This will create a file called `.httr-oauth` in your current directory. NEVER CHECK THIS INTO GIT. (In the RStudio git pane, right-click on it and select ignore, then press OK.)

```{r}
gs_key()
```
