---
unit_title: "Other single table verbs"
theme: wrangle
needs: [manip-basics]
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

You've learned the most important verbs for data analysis: `filter()`, `mutate()`, `group_by()` and `summarise()`. There are a number of other verbs that are not quite as important but still come in handy from time-to-time. The goal of this document is to familiarise you with their purpose and basic operation

```{r setup, message = FALSE}
library(tidyverse)
library(nycflights13)
```

## Select

Most of the datasets you'll work with in this class only have a relatively small number of variables, and generally you don't need to reduce further. In real life, you'll sometimes encounter datasets with hundreds or even thousands of variables, and the first challenge is just to narrow down to a useful subset. Solving that problem is the job of `select()`.

`select()` allows you to work with column names using a handful of helper functions:

* `starts_with("x")` and `ends_with("x")` select variables that start with a
  common prefix or end with a common suffix.
  
* `contains("x")` selects variables that contain a phrase. `matches("x.y")`
  select all variables that match a given regular expression (which you'll 
  learn about later in the course).
  
* `a:e` selects all variables from variable `a` to variable `e` inclsive.

You can also select a single varible just by using its name directly.

```{r}
flights %>% select(year:day, ends_with("delay"))
```

To remove variables from selection, put a `-` in front of the expression.

```{r}
flights %>% select(-starts_with("dep"))
```

There's one last helper that's useful if you just want to move a few variables to the start: `everything()`.

```{r}
flights %>% select(dep_delay, arr_delay, everything())
```

## Rename

To change the name of a variable use `df %>% rename(new_name = old_name)`. If you have trouble remembering which sides old and new go on, remember it's the same order as `mutate()`.

```{r}
flights %>% rename(tail_num = tailnum)
```

If you're selecting and renaming, note that you can also use `select()` to rename. This sometimes allows you to save a step.

```{r}
flights %>% select(year:day, tail_num = tailnum)
```


## Transmute

Transmute is a minor variation of `mutate()`. The main difference is that it drops any variables that you didn't explicitly mention. It's a useful shortcut for `mutate()` + `select()`.

```{r}
df <- tibble(x = 1:3, y = 3:1)

# mutate() keeps all the variables
df %>% mutate(z = x + y)

# transmute() drops all the variables
df %>% transmute(z = x + y)
```

## Arrange

`arrange()` lets you change the order of the rows. To put a column in descending order, use `desc()`.

```{r}
flights %>% arrange(desc(dep_delay))
flights %>% arrange(year, month, day)
```

## Distinct

`distinct()` removes duplicates from a dataset. The result is ordered by first occurence in original dataset.

```{r}
flights %>% distinct(carrier, flight)
```


## Sample

When working with very large datasets, sometimes it's convenient to reduce to a smaller dataset, just by taking a random sample. That's the job of `sample_n()` and `sample_frac()`. `sample_n()` selects the same number of observations from each group, `sample_frac()` selects the same proportion.

```{r}
popular_dest <- flights %>%
  group_by(dest) %>%
  filter(n() > 1000)

# Creates a dataset with the same number of flights to each dest
popular_dest %>% sample_n(100)

# Creates a dataset with proportion number of flights to each dest
popular_dest %>% sample_frac(0.01)
```

## Slice rows

`slice()` allows to pick rows by position, by group. `head()` and `tail()` just show the first (or last) few rows of the entire data frame. 

```{r}
# first flights to each dest
flights %>% group_by(dest) %>% slice(1:5)

# first flights overall
flights %>% head()

# last flights overall
flights %>% tail()
```

