---
unit_title: Scoped verbs
needs: [iteration, manip-basics]
theme: wrangle
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(tibble.print_min = 5, tibble.print_max = 5)
library(tidyverse)
```

## dplyr 0.6.0

First, make sure you have the latest version of dplyr, 0.6.0. You can check what version you currently have with `packageVersion()`:

```{r}
packageVersion("dplyr")
```

(If you're using the development version, version 0.5.0.9005 is also ok.)

If you're version is older, use `install.packages()` to update it.

## Scoped verbs

In the latest version of dplyr each of the single table verbs comes in three additional forms with the suffixes `_if`, `_at`, and `_all`. These __scoped__ variants allow you to affect multiple variables at once:

* `_if` allows you to pick variables based on a predicate function like
  `is.numeric()` or `is.character()`.
  
* `_at` allows you to pick variables using the same syntax as `select()`.

* `_all` operates on all variables.

I'll illustrate the three variants in detail for `summarise()`, then explore in less detail how you can use similar techniques with `mutate()` and `filter()`. You'll need the scoped variants of the other verbs less frequently, but when you do, it should be straightforward to generalise what you've learn here.

## Summarise

### `summarise_all()`

The simplest variant to understand is `summarise_all()`. It takes a tibble and a function and applies that function to each column:

```{r}
df <- tibble(
  x = runif(100),
  y = runif(100),
  z = runif(100)
)
summarise_all(df, mean)
```

If you want to apply multiple summaries, use the `funs()` helper:

```{r}
summarise_all(df, funs(min, max))
```

There are two slightly inconsistent ways to use an inline function (this inconistency will get ironed out in a future version).

```{r}
# For a single function, use ~ 
summarise_all(df, ~ sd(.) / mean(.))

# For multiple functions, use funs(), dropping the ~
# Typically you'll want to name the function so you get reasonable
# variable names.
summarise_all(df, funs(cv = sd(.) / mean(.), mean))
```

### `summarise_at()`

`summarise_at()` allows you to pick columns in the same way as `select()`, that is, based on their names. There is one small difference: you need to wrap the complete selection with the `vars()` helper (this avoids ambiguity).

```{r}
summarise_at(df, vars(-z), mean)
```

By default, the newly created columns have the shortest names needed to uniquely identify the output.

```{r}
summarise_at(df, vars(x), funs(min, max))
summarise_at(df, vars(x, y), min)
summarise_at(df, vars(-z), funs(min, max))
```

See the examples in the documentation if you want to force names when they're not otherwise needed.

### `summarise_if()`

`summarise_at()` allows you to pick variables to summarise based on their name. `summarise_if()` allows you to pick variables to summarise based on some property of the column. Typically this is their type because you want to (e.g.) apply a numeric summary function only to numeric columns:

```{r}
starwars %>%
  group_by(species) %>%
  summarise_if(is.numeric, mean, na.rm = TRUE)
```

(Note that `na.rm = TRUE` is passed on to `mean()` in the same way as in `purrr::map()`.)

## Mutate

`mutate_all()`, `mutate_if()` and `mutate_at()` work in a similar way to their summarise equivalents.

```{r}
mutate_all(df, log10)
```

Often you'll want to use an inline expression. As above, either use `~` for a single function or `funs()` for multiple functions:

```{r}
mutate_all(df, ~ round(. * 25))
mutate_all(df, funs(half = . / 2, double = . * 2))
```

The default names are generated in the same way as `summarise()`. That means that you may want to use a `transmute()` variant if you want to apply multiple transformations and don't want the original values:

```{r}
transmute_all(df, funs(half = . / 2, double = . * 2))
```

## Filter

`filter_all()` is the most useful of the three `filter()` variants. You use it conjunction with `all_vars()` or `any_vars()` depending on whether or not you want rows where all variables meet the criterion, or where just one variable meets it.

It's particularly useful finding missing values:

```{r}
library(nycflights13)

# Rows where any value is missing
filter_all(weather, any_vars(is.na(.)))

# Rows where all wind variables are missing
filter_at(weather, vars(starts_with("wind")), all_vars(is.na(.)))
```

