---
unit_title: Parsing basics
needs: [data-structure-basics, manip-basics]
theme: wrangle
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

So far you've worked with data sets that have been bundled in R packages, or have been created with `tibble()` or `tribble()`. Now it's time to learn how to read simple flat files from disk. To do this, we'll use functions from [readr](http://readr.tidyverse.org). readr is one of the core tidyverse packages, so you won't usually load it explicitly.

```{r}
library(tidyverse)
```

## Delimited files

In this unit, we're going to focus on delimited files. Delimited files have a __delimiter__ between each value. Two types make up the majority of delimited files that you'll see in the wild: csv (comma separated) and tsv (tab separated). We'll focus on csv files, but everything you'll learn applies equally to tsvs, replacing commas with tabs.

A typical csv file looks something like this:

```
Sepal.Length,Sepal.Width,Petal.Length,Petal.Width,Species
5.1,3.5,1.4,0.2,setosa
4.9,3,1.4,0.2,setosa
4.7,3.2,1.3,0.2,setosa
4.6,3.1,1.5,0.2,setosa
5,3.6,1.4,0.2,setosa
5.4,3.9,1.7,0.4,setosa
4.6,3.4,1.4,0.3,setosa
5,3.4,1.5,0.2,setosa
```

Note that:

* The first line gives the column names
* Each subsequent line is one row of data
* Each value is separated by a comma (hence the name)

Typically you can recognise a csv file by its extension: `.csv`. But beware! Sometimes the extension lies, and if you're getting weird errors when reading a file, it's a good idea to peek inside the file using `readr::read_lines()` and `writeLines()`, specifying the `n_max` argument to just look at the first few lines. (You'll learn more about `writeLines()` when we get to strings; for now just remember it's a useful tool for printing lines to the screen.)

```{r}
"heights.csv" %>% 
  read_lines(n_max = 10) %>%
  writeLines()
```

This file illustrates another feature present in many csv files: some values are surrounded by quotes. Confusingly, this isn't a guarantee that the value is a string: some csv files also surround numbers in quotes too.  As you work with more csv files you'll discover there are few hard and fast rules: for pretty much every crazy thing that you can imagine, someone has done it in a csv file somewhere.

## `read_csv()`

The workhorse for reading in csv files is called `read_csv()`. You give it a path to a csv file and it gives you back a tibble:

```{r}
heights <- read_csv("heights.csv")
heights
```

If you are very lucky, you can point `read_csv()` at a file and it just works. But this is usually the exception, not the rule, and often you'll need to tweak some arguments.

The most important arguments to `read_csv()` are:

*   `col_names`: usually `col_names = TRUE` which tells `read_csv()` that
    the first line of the file is the column names. If there aren't any
    column names set `col_names = FALSE` or supply a character vector telling
    `read_csv()` what they should be `col_names = c("x", "y", "z")`
    
*   `col_types`: you might have noticed that when we called `read_csv()` above
    it printed out a list of column "specifications". That describes how readr
    converts each column into an data structure. readr uses some pretty good 
    heuristics to guess the type, but sometimes the heuristics fail and you'll
    need to supply the truth. You'll learn more about that later in the course
    
*   It's fairly common to encounter csv files that have a bunch of 
    `r "\U0001f4a9"` at the top. You can use `skip = n` to skip the first n 
    lines, or `comment = "#"` to ignore all lines that start with `#`.

*   `read_csv()` expects missing values to be suppled as `NA`. If your file
    uses a different convention, use `na = "."` to override the default.

You'll get to practice using these arguments in the exercises.
