---
unit_title: "Spatial basics"
theme: wrangle
needs: [list-cols]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Spatial packages

In R, there are two main lineages of tools for dealing with spatial data: sp and sf. 

* sp has been around for a while (the first release was in 2005), and it has a
  rich ecosystem of tools built on top of it. However, it uses a rather 
  data structure, which can make it challenging to use. 
  
* sf is newer (first released in October 2016!) so it doesn't have such a 
  rich ecosystem. However, it's much easier to use and fits in very naturally 
  with the tidyverse, and the ecosystem around it will grow rapidly.
   
In this class, we're going to use sf, so start by installing it:

```{r, eval = FALSE}
install.packages("sf")
```

## Loading data

To read spatial data in R, use `read_sf()`. The following example reads an example dataset built into the sf package:

```{r, message = FALSE}
library(tidyverse)
library(sf)

# The counties of North Carolina
nc <- read_sf(system.file("shape/nc.shp", package = "sf"), 
  quiet = TRUE,  
  stringsAsFactors = FALSE
)
```

I recommend always setting `quiet = TRUE` and `stringsAsFactors = FALSE`. 

Here we're loading from a __shapefile__ which is the way spatial data is most commonly stored. Despite the name a shapefile isn't just one file, but is a collection of files that have the same name, but different extensions. Typically you'll have four files:

* `.shp` contains the geometry, and `.shx` contains an index into that geometry.

* `.dbf` contains metadata about each geometry (the other columns in the data
  frame).

* `.prf` contains the coordinate system and projection information. You'll 
  learn more about that shortly.

`read_sf()` can read in the majority of spatial file formats, so don't worry if your data isn't in a shapefile; the chances are `read_sf()` will still be able to read it.

## Converting data

If you get a spatial object created by another package, us `st_as_sf()` to convert it to sf. For example, you can take data from the maps package (included in base R) and convert it to sf:

```{r}
library(maps)
nz_map <- map("nz", plot = FALSE)
nz_sf <- st_as_sf(nz_map)
```

## Data structure

`nc` is a data frame, and not a tibble, so when printing, it's a good idea to use `head()` so you only see the first few rows:

```{r}
head(nc)
head(nz_sf)
```

This is an ordinary data frame, with one exception: the __geometry__ column. This column contains simple "features", a standard way of representing two dimesional geometries like points, lines, polygons, multilines, and multipolygons. Multilines and multipolygons are nededed to represent geographic phenomena like a river with multiple branches, or a state made up of multiple islands.

```{r}
nc$geometry
```

Use `plot()` to show the geometry. You'll learn how to use ggplot2 for more complex data visualisations in the next unit.

```{r nc-plot}
plot(nc$geometry)
```

## Manipulating with dplyr

Since an sf object is just a data frame, you can manipulate it with dplyr. The following example gives you a taste:

```{r}
nz_sf %>%
  mutate(area = as.numeric(st_area(geometry))) %>%
  filter(area > 1e10)
```

`st_area()` returns an object with units (i.e. $m^2$), which is precise, but a little annoying to work with. I used `as.numeric()` to convert to a regular numeric vector.

## Geometry

The geometry column is a list-column. You'll learn more about list-columns later in the course, but in brief, they're the richest and most complex type of column because a list can contain any other data structure including other lists.

It's worthwhile to pull out one piece so you can see what's going on under the hood:

```{r}
str(nc$geometry[[1]])
plot(nc$geometry[[1]])
```

Note the use of `[[` to extract a single element, here, the first polygon.

This is list of lists of matrices:

*   The top-level list has one element for each "landmass" in the county. We
    can find a more interesting case:
  
    ```{r}
    n <- nc$geometry %>% map_int(length)
    table(n)
    
    interesting <- nc$geometry[n == 3][[1]]
    plot(interesting)
    str(interesting)
    ```
    This is a county made up of three non-contiguous pieces.

*   The second-level list is not used in this dataset, but is needed when 
    you have a landmass that contains an lake. (Or a landmass that contains
    an lake which has an island which has a pond).
    
*   Each row of the matrix gives the location of a point on the boundary of 
    the polygon.

## Coordinate system

To correctly plot spatial data, you need know exactly what the numeric positions mean, i.e. what are they in reference to? This is called the __coordinate reference system__ or CRS. Often spatial data is described in terms of latitude and longitude. You can check this with `st_is_longlat()`:

```{r}
st_is_longlat(nc)
```

You might think that if you know the latitude and longitude of a point, you know exactly where it is on the Earth. However, things are not quite so simple, because latitude and longitude are based on the assumption that the Earth is a smooth ellipsoid, which is not true. Because different approximations work better in differently places, most countries have their own approximation: this is called the __geodetic datum__, or just __datum__ for short.

Take two minutes and watch this simple explanation of the datum: https://www.youtube.com/watch?v=xKGlMp__jog

To get the datum and other coordinate system metadata, use `st_crs()`:

```{r}
st_crs(nc)
```

Here the datum is "NAD27", the [North American Datum](https://en.wikipedia.org/wiki/North_American_Datum) of 1927 (NAD27)

In this class, you won't have to worry too much about the datum as sf and ggplot2 will take care of the details for you. But it's good to know why it exists and how to identify it if something goes wrong.
