```{r setup, include = FALSE}
library(tidyverse)
```

1.  It's easier to see the relationship between `price` and `carat` if 
    you if log transform both variables.  Compare and contrast the 
    following three approaches. What are the advantages and disadvantages?
    
    ```{r, eval = FALSE}
    ggplot(diamonds, aes(log10(carat), log10(price))) +
      geom_hex()
    
    ggplot(diamonds, aes(carat, price)) +
      geom_hex() + 
      scale_x_log10() + 
      scale_y_log10()
    
    diamonds %>% 
      mutate(lcarat = log10(carat), lprice = log10(price)) %>% 
      ggplot(aes(lcarat, lprice)) +
      geom_hex()
    ```
    
1.  The following plot shows how the delay varies by time of day
    over the course of the year:
    
    ```{r}
    library(nycflights13)
    
    flights %>% 
      group_by(
        month = lubridate::month(time_hour, label = TRUE), 
        hour = lubridate::wday(hour)
      ) %>% 
      summarise(delay = mean(dep_delay, na.rm = TRUE)) %>% 
      ggplot(aes(hour, month)) + 
        geom_tile(aes(fill = delay))
    ```

    Improve this plot by:
    
    1. Choosing a better colour scale
    1. Setting x limits focus on where most of the data lies.
    1. Flipping the direction of the y axis (so earliest is at the top)
